Adaptive Waveform Inversion for Transmitted Wave Data

### 2024-11-08 repro/papers/awi/paper.tex
more cleanup

Suppose conversely that $\|\Sl[\kappa]\ul[\kappa;d] - \dl\| \le
\frac{1}{2}\|d\|$. Then $\|\Sl[\kappa]\ul[\kappa;d] \| \ge
\frac{1}{2}\|d\|$, so
\[
  \|d\| \le 2\|\Fl[\kappa]*\ul[\kappa;d] \le
  2\lambda^{\frac{1}{2}}\|\ul[\kappa;d]\|
\]
from Young's inequality, Proposition \ref{thm:noisefree}, and
inequalities \ref{\eqn:lambdanorms}. From the bound \ref{eqn:unorm} this is in
turn
\[
  \le R^* \frac{\lambda}{\sigma} \|d\|
\]
whence
\[
  \sigma \le R^* \lambda.
\]
for a suitable $R^*>0$ independent of $\log \kappa \in M$ and $\lambda
\in (0,1]$. This establishes conclusion 2a in Theorem \ref{thm:main}, and
suggests that in order to expect filtered predicted data
$\Sl[\kappa]\ul[\kappa;d\|$ to remian uniformly (in $\lambda$) close to the data to be
inverted, as the heuristic justification for AWI suggests, the
regularization weight $\sigma$ must be coupled to the wavelength proxy
$\lambda$. Therefore we assume from now on that equation
\ref{eqn:lamsig} holds,  
in which $r$ is a positive constant with dimensions pressure$^2 \times 
$ time$^2$.

@@

\begin{proof} {\em of Theorem \ref{thm:main}}:
Conclusions 1 and 2a have already been established. Conclusion 2b
follows from equation \ref{eqn:Sutz} in Proposition \ref{thm:leading},
combined with the first estimate in \ref{eqn:remest} and the bounds
\ref{eqn:Srembd}.

@@

\begin{prop}
  \label{thm:MSWIepsalphalim}
Suppose that $\kappa^* \in M$ and $\dl=\Fl[\kappa^*]$.
\begin{equation}
  \label{eqn:eq7}
  \lim_{\epsilon \rightarrow 0}\lim_{\alpha \rightarrow 0}
  \frac{1}{\alpha^2}\left(\tJase[\kappa;\dl]-\tilde{J}_{\sigma}[\kappa;d]\right)=
 \|T\ul\|^2 =  J_{\rm MSWI,\sigma}[\kappa;d]
\end{equation}
in which $\ul$ is the solution of the least squares problem
\ref{eqn:reg} with $S=\Sl$, $d=\dl$, and $J_{\rm MSWI,\sigma}$ is as
defined in equation \ref{eqn:mswidef}.
\end{prop}

@@

\noindent {\bf Remark:} From inequalities \ref{eqn:ulnorm},
\ref{eqn:Srembd}, $\tilde{J}_{r\lambda}[\kappa;\dl] \le C$. That
suggests that the contribution of the second term on the
left-hand-side of the limit \ref{eqn:cor2} is insignificant for small
$\lambda$, whereas it is comparable to the first term in
\ref{eqn:cor3}.

@@

\section{Multiple Arrivals and Cycle-Skipping}
The geometric asymptotics approximation \ref{eqn:fwdgreen} holds only
in the ``single arrival'' case, that is, when sources and receivers
are close enough together relative to the length scale of significant
variation in the wave velocity $((\kappa\beta)^{1/2})$ that a unique ray of
geometric acoustics connects each source-receiver pair. At larger
distances, multiple raypaths typically exist, corresponding to
multiple idenfiable wavefront arrivals \cite[]{Whi:82}. Accordingly, in such settings
the analysis presented in the last section no longer applies, and the
relation established there between AWI and travel-time inversion is
case into doubt.

In fact, the behaviour inherited from travel-time inversion in the
single arrival case, namely the absence of spurious stationary points
(``cycle-skipping''), generally fails when multiple arrivals with
significant energy are present in the data. \cite{Symes:94c}
illustrated this phenomenon for a version of MSWI. Crosswell
source-receiver geometry often generates multiple arrivals,
particularly near low-velocity sedimentary layers that are natural
targets of seismic exploration and act as wave
guides. \cite{Plessix:00} successfully applied a version of MSWI to
crosswell waveform inversion by removing traces with close source and
receiver depths, thus leaving only data that tended to be dominated by
single arrivals. \cite{HuangSymes:Geo17} and
\cite{SymesChenMinkoff:24} illustrate the failure of the
version of MSWI reported here to avoid cycle-skipping in inversion of
strongly refracting models, using frequency-domain implementation of
the basic definition \ref{eqn:mswidef} and time-domain implementation
of the reduced penalty function \ref{eqn:mswipenred} respectively. \cite{Yongetal:GJI23}
have shown that precisely the same issue afflicts AWI, and have
proposed a remedy by localizing the definition of the objective to
short time intervals, within which multiple wavefront arrrivals are
unlikely. \cite{Symes:23} constructs an extension based on surface
sources, as opposed to the source-receiver extension which is the
basis of MSWI and AWI. This surface source extension generates an
invertible map from extended source to data.

@@

Of courese, the adaptive filter just computed is not square
integrable, and must be regularized for use in a square-integral
construction like that of AWI or MSWI. That amounts essentially to
replacing the $\delta$s in the above formulae by a square-integrable
approximate Dirac function, of roughly a wavelength in width. In the
preceding section, this approximate Dirac was $g$, so use the same
notation here. The regularized version of $u(t)$ (solution of the
least squares problem \ref{eqn:reg}) contains summands with multiples of $g(t-(\tau^*_1-\tau_0))$
and $g(t-(\tau_1-\tau^*_0))$. These are approximately scaled by
$\tau^*_1-\tau_0$ and $\tau_1-\tau^*_0$ respectively in the
definitions \ref{eqn:mswidef} and \ref{eqn:awidefbis} of the MSWI and AWI
objectives. For other choices of objective, they contribute similarly
to the total. They will eventually begin to cancel as $m$ approaches
$\kappa^*$, however this cancellation happens only when the times are
within the width of the approximate 
Dirac waveform $g$ - that is, within a wavelength.

@@

These constraints are discussed in detail in the next section. Some
could certainly be relaxed. For instance, exponential decay of local
energy implies that the finite time duration of actual measured data
can be treated as a perturbation of the infinite time data setting
posed here. The exception is the single-arrival hypothesis: it is
essential, as without it, the link between the travel time and
extended waveform inversion is lost. This is a feature of the
source-receiver extension, not of a particular objective function
(MSWI, AWI,...) formulated in terms of it. This extension treats each receiver signal
independently. If only a single identifiable waveform arrives at
each receiver, AWI (and similar approaches) can implicitly identify
its arrival time, and the link to travel time inversion is
established. If distinct waveforms are present at multiple times,
the single receiver signal does not determine which waveform in the
predicted signal should align with which waveform in the data to be
fitted. This ``crosstalk'' ambiguity is analyzed in the fifth
section, and afflicts all approaches based on the source-receivre
extension.

In any case, these hypotheses enable a simple and clear
analysis of AWI, and seem a reasonable starting point for further
investigation.

In this paper, I have shown that AWI is a good approximation to least
squares travel time inversion of transmitted wave data, up to errors proportional to a typical
data wavelength and to a regularization (``pre-whitening'')
weight. However, the argument leading to this conclusion works only if
the material model is assumed slowly varying so that geometric
acoustics accurately describes the solution of the wave equation, and if
sources and receivers are sufficiently close to one another to be
connected by unique rays. The first condition may be ignored to some
extent - the conclusion of this paper likely holds if the data is
dominated by ballistic transmitted waves. The second condition
is essential: even for purely transmitted waves, the presence of
multiple arrivals breaks the connection
with travel time inversion, and AWI and similar algorithms
become essentially as prone to cycle-skipping as is FWI.

%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%

\subsection{AWI via Optimization}

AWI is based an {\em extension} of the modeling operator $F$, that is,
an operator ${\bar F}$ with larger domain, coinciding with $F$ on its domain..
The extended modeling operator ${\bar F}$ maps extended models
$\bar{m}=(\kappa,\bar{f})$ incorporating extended sources
$\bar{f}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure
field. Denote by $\bar{p}(\bx,t;\bx_s,\bx_r), \bar{\bv}(\bx,t;\bx_s,\bx_r)$
the solution of the acoustic system \ref{eqn:awe} with $f(t)$ in the
right-hand side of the first equation replaced by
$\bar{f}(\bx_r,\cdot;\bx_s)$. Then
$\bar{F}[\kappa,\bar{f}](\bx_r,t;\bx_s) =
\bar{p}(\bx_r,t;\bx_s,\bx_r)$. As was the case for $F$, the Hadamard
decomposition \ref{eqn:green} and conditions G1-G4, W1-W3 imply that
$\bar{F}[\kappa, \cdot]$ is a bounded map: $L^2(X_{sr} \times \bR)
\rightarrow L^2(X_{sr} \times \bR)$. If all  
of the wavelets are the same, that is, ${\bar f}({\bf  
  x}_r,t;{\bf x}_s) = f(t)$ is independent of source and receiver  
position, then ${\bar F}[\kappa,\bar{f}] = F[\kappa]$. That is, $F$ is a special
case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of
$F$. This is the {\em source-receiver} extension, in the terminology of \cite{HuangSymes2015SEG}.

\cite{Warner:16} represent the extended sources $\bar{f}$ as time convolutions of the
(known) source $f$ with an {\em adaptive filter} $u \in L^2(X_{sr}
\times \bR)$, that is, $\bar{f}=u * f$.

\section{BLAH BLAH}

Assume as in the last section, that is, that solutions
of the acoustic system \ref{eqn:awe} is well-approximated by the
geometric asymptotics as in equation \ref{eqn:hadamard}, and that the
data is noise-free ($d=F[\kappa^*]$). Together with the properties of $w$
just listed, these assumptions imply that
\begin{equation}
  \label{eqn:eq11}
  J_{0,\sigma}[\kappa;d] = O(\sigma) + O(\lambda(w)).
\end{equation}

\begin{proof}
  Since the arguments of $\Sl$,... are just along for the ride
  suppress them, as before.
  First analyse the leading order approximation: from equations
  \ref{eqn:hatuf} and \ref{eqn:hatgf},
  \[
  \|\Sla \uzl -\dzl\|^2 = \frac{(a^*)^2}{2\pi} \int d\omega \left| \fwl(\omega) \frac{|\fwl(\omega)|^2}{|\fwl(\omega)|^2 + \frac{r\lambda}{a^2}}
    - \fwl(\omega)\right|^2
\]
\[
  =r^2\lambda^2\left(\frac{(a^*)^2}{2\pi a^4}\right) \int d\omega \frac{| \fwl(\omega)|^2}
  {(|\fwl(\omega)|^2 + \frac{r\lambda}{a^2})^2}
\]
\[
  =r^2\lambda^2\left(\frac{(a^*)^2}{2\pi a^4}\right) \int d\omega
  \frac{| \lambda^{\frac{1}{2}}w_1(\lambda \omega)|^2}
  {(|\lambda^{\frac{1}{2}}w_1(\lambda \omega)|^2 + \frac{r\lambda}{a^2})^2}
\]
\[
  = \frac{1}{2\pi}\left(\frac{ra^*}{a}\right)^2 \int d\omega
  \frac{|aw_1(\omega)|^2}
  {(|aw_1(\omega)|^2 + r)^2}
\]
which depends on $M$ (and $r$) but not on $\lambda$. From the result
\ref{eqn:u0snormlam}, the same is true of the second term in the
objective function defined in the third equation  of the display
\ref{eqn:leading}: that is,
\begin{equation}
  \label{eqn:leadingconst}
  \|\Sla \uzl -\dzl\|^2 + r\lambda \|\uzl\|^2 = C
\end{equation}

\end{proof}
To see this, begin with the definition:
\[
  \tilde{J}_{0,\sigma}[\kappa;d] = \frac{1}{2}
  (\|S_{\lambda}[\kappa]u_0[\kappa_{\lambda};d]-d\|^2 + \lambda \|u_0[\kappa_{\lambda};d]\|^2.
\]
Using the approximate identities \ref{eqn:hatuf} and \ref{eqn:hatgf} and
rearranging,
\[
\|\Sl[\kappa]u_0[\kappa_;d]-d\|^2 
  \approx \int d\omega
  |a\hat{w}\hat{u_0[\kappa_{\lambda};d]}e^{i\omega\tau}-a^*\hat{w}e^{i\omega\tau^*}|^2
\]
\[
  = \int d\omega
  (a^*)^2|\hat{w}|^2\left(\frac{\lambda}{a^2|\hat{w}|^2 +
      \lambda}\right)^2   
\le \int d\omega (a^*)^2|\hat{w}|^2
  \left(\frac{\lambda}{a^2|\hat{w}|^2 + \lambda}\right)
\]
\begin{equation}
  \label{eqn:eq12}
  = \lim_{\Omega \rightarrow \infty} \int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2\lambda
  \left(\frac{a^2|\hat{w}|^2}{a^2|\hat{w}|^2 + \lambda}\right)
\end{equation}
For any $\Omega > 0$,
\[
\int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2\lambda\left(\frac{a^2|\hat{w}|^2}{a^2|\hat{w}|^2 + \lambda}\right)
\le \int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2 \lambda
  \left(\frac{ a|\hat{w}|}{(a^2|\hat{w}|^2 +
      \lambda)^{\frac{1}{2}}}\right)
\]
\[
  = \sigma \int_{-\Omega}^{\Omega} d\omega \left(\frac{a^*}{a}\right)^2 a|\hat{w}|
  \left(\frac{\sigma }{(a^2|\hat{w}|^2 +\lambda)^{\frac{1}{2}}}\right)
  \le \sigma \int_{-\Omega}^{\Omega} d\omega \frac{(a^*)^2}{a}
  |\hat{w}|
  \le \sigma \frac{(a^*)^2}{a} \|\hat{w}\|_{L^1(\bR)}.
\]
Together with the approximate inequality \ref{eqn:eq12}, this
inequality implies
that
\[
  \|S_{\lambda}[\kappa]u_0[\kappa_{\lambda};d] - d\|^2 = O(\sigma) + O(\lambda(w)).
\]
A similar argument leads to the conclusion that
\[
  \lambda\|u_0[\kappa_{\lambda};d]\|^2 = O(\sigma) + O(\lambda(w)).
\]
Together, these conclusions imply the assertion \ref{eqn:eq11}.

%%%%%%%%%%%%%%%%%%%%

>>>>

Suppose conversely that $\|\Sl[\kappa]\ul[\kappa;d] - \dl\| \le
\frac{1}{2}\|d\|$. Then $\|\Sl[\kappa]\ul[\kappa;d] \| \ge
\frac{1}{2}\|d\|$, so
\[
  \|d\| \le 2\|\Fl[\kappa]*\ul[\kappa;d] \le
  2\lambda^{\frac{1}{2}}\|\ul[\kappa;d]\|
\]
from Young's inequality, Proposition \ref{thm:noisefree}, and
inequalities \ref{eqn:lambdanorms}. From the bound \ref{eqn:unorm} this is in
turn
\[
  \le R^* \frac{\lambda}{\sigma} \|d\|
\]
whence
\[
  \sigma \le R^* \lambda.
\]
for a suitable $R^*>0$ independent of $\log \kappa \in M$ and $\lambda
\in (0,1]$. This establishes conclusion 2a in Theorem \ref{thm:main}, and
suggests that in order to expect filtered predicted data
$\Sl[\kappa]\ul[\kappa;d\|$ to remian uniformly (in $\lambda$) close to the data to be
inverted, as the heuristic justification for AWI suggests, the
regularization weight $\sigma$ must be coupled to the wavelength proxy
$\lambda$. Therefore we assume from now on that equation
\ref{eqn:lamsig} holds,  
in which $r$ is a positive constant with dimensions pressure$^2 \times 
$ time$^2$.

@@

\begin{proof} {\em of Theorem \ref{thm:main}}:
Conclusions 1 and 2a have already been established. Conclusion 2b
follows from equation \ref{eqn:Sutz} in Proposition \ref{thm:leading},
combined with the first estimate in \ref{thm:remest} and the bounds
\ref{eqn:Srembd}.

@@

\begin{prop}
  \label{thm:MSWIepsalphalim}
Suppose that $\kappa^* \in M$ and $\dl=\Fl[\kappa^*]$.
\begin{equation}
  \label{eqn:eq7}
  \lim_{\epsilon \rightarrow 0}\lim_{\alpha \rightarrow 0}
  \frac{1}{\alpha^2}\left(\tJase[\kappa;\dl]-\tilde{J}_{\sigma}[\kappa;d]\right)=
 \|T\ul\|^2 =  J_{\rm MSWI,\sigma}[\kappa;d]
\end{equation}
in which $\ul$ is the solution of the least squares problem
\ref{eqn:reg} with $S=\Sl$, $d=\dl$, and $J_{\rm MSWI,\sigma}$ is as
defined in equation \ref{eqn:mswidefmain}.
\end{prop}

@@

\noindent {\bf Remark:} From inequalities \ref{eqn:ulnorm},
\ref{eqn:Srembd}, $\tilde{J}_{r\lambda}[\kappa;\dl] \le C$. That
suggests that the contribution of the second term on the
left-hand-side of the limit \ref{eqn:cor1mswi} is insignificant for small
$\lambda$, whereas it is comparable to the first term in
\ref{eqn:cor3}.

@@

\section{Multiple Arrivals and Cycle-Skipping}
The geometric asymptotics approximation \ref{eqn:green} holds only
in the ``single arrival'' case, that is, when sources and receivers
are close enough together relative to the length scale of significant
variation in the wave velocity $((\kappa\beta)^{1/2})$ that a unique ray of
geometric acoustics connects each source-receiver pair. At larger
distances, multiple raypaths typically exist, corresponding to
multiple idenfiable wavefront arrivals \cite[]{Whi:82}. Accordingly, in such settings
the analysis presented in the last section no longer applies, and the
relation established there between AWI and travel-time inversion is
case into doubt.

In fact, the behaviour inherited from travel-time inversion in the
single arrival case, namely the absence of spurious stationary points
(``cycle-skipping''), generally fails when multiple arrivals with
significant energy are present in the data. \cite{Symes:94c}
illustrated this phenomenon for a version of MSWI. Crosswell
source-receiver geometry often generates multiple arrivals,
particularly near low-velocity sedimentary layers that are natural
targets of seismic exploration and act as wave
guides. \cite{Plessix:00} successfully applied a version of MSWI to
crosswell waveform inversion by removing traces with close source and
receiver depths, thus leaving only data that tended to be dominated by
single arrivals. \cite{HuangSymes:Geo17} and
\cite{SymesChenMinkoff:24} illustrate the failure of the
version of MSWI reported here to avoid cycle-skipping in inversion of
strongly refracting models, using frequency-domain implementation of
the basic definition \ref{eqn:mswidefmain} and time-domain implementation
of the reduced penalty function \ref{eqn:mswipenred} respectively. \cite{Yongetal:GJI23}
have shown that precisely the same issue afflicts AWI, and have
proposed a remedy by localizing the definition of the objective to
short time intervals, within which multiple wavefront arrrivals are
unlikely. \cite{Symes:23} constructs an extension based on surface
sources, as opposed to the source-receiver extension which is the
basis of MSWI and AWI. This surface source extension generates an
invertible map from extended source to data.

@@

Of courese, the adaptive filter just computed is not square
integrable, and must be regularized for use in a square-integral
construction like that of AWI or MSWI. That amounts essentially to
replacing the $\delta$s in the above formulae by a square-integrable
approximate Dirac function, of roughly a wavelength in width. In the
preceding section, this approximate Dirac was $g$, so use the same
notation here. The regularized version of $u(t)$ (solution of the
least squares problem \ref{eqn:reg}) contains summands with multiples of $g(t-(\tau^*_1-\tau_0))$
and $g(t-(\tau_1-\tau^*_0))$. These are approximately scaled by
$\tau^*_1-\tau_0$ and $\tau_1-\tau^*_0$ respectively in the
definitions \ref{eqn:mswidefmain} and \ref{eqn:awidefmain} of the MSWI and AWI
objectives. For other choices of objective, they contribute similarly
to the total. They will eventually begin to cancel as $m$ approaches
$\kappa^*$, however this cancellation happens only when the times are
within the width of the approximate 
Dirac waveform $g$ - that is, within a wavelength.

@@

Some of these constraints
could certainly be relaxed. For instance, exponential decay of local
energy implies that the finite time duration of actual measured data
can be treated as a perturbation of the infinite time data setting
posed here. The conclusions of this paper likely hold in some form
even for non-smooth sets of models, provided that the data for such models is
dominated by ballistic transmitted waves. The exception is the single-arrival hypothesis: it is
essential, as without it, the link between the travel time and
extended waveform inversion is lost. This is a feature of the
source-receiver extension, not of a particular objective function
(MSWI, AWI,...) formulated in terms of it. This extension treats each receiver signal
independently. If only a single identifiable waveform arrives at
each receiver, AWI (and similar approaches) can implicitly identify
its arrival time, and the link to travel time inversion is
established. If distinct waveforms are present at multiple times,
the single receiver signal does not determine which waveform in the
predicted signal should align with which waveform in the data to be
fitted. This ``crosstalk'' ambiguity is analyzed in the fifth
section, and afflicts all approaches based on the source-receivre
extension.

@@
